### 2024-11-21 paper/background.tex
Write background

\red{TODO: fix fig name}

@@

% breif overview
WebAssembly~\cite{wasm} is a low-level bytecode language that is safe, fast,
portable, and compact.
It is widely used as a compile target for web applications.
Not only that, other areas such as edge computing~\cite{wasm-edge},
IoT~\cite{wasm-iot}, and block chains~\cite{wasm-block} deploy the advantages
of WebAssembly.

% risk of implementation divergence
There are dozens of WebAssembly engines; all the browsers have there own
implementations of WebAssembly with multiple tiers ~\cite{v8}
\cite{spidermonkey} \cite{webkit}, and there are many engines that target for
specific areas including embedded systems and edge computing.
However, as WebAssembly should be portable across these implementations, it risks
the implementations divergence.

% rigorous standardization -> spec is rigorous: formal notation & prose notation
To mitigate this problem, WebAssembly has been standardized very rigourously by
the W3C~\cite{wasm-w3c}.
Especially, WebAssembly specification document is written very rigorously.
It describes the semantics of WebAssembly in two forms: formal notation and
prose notation.
Formal notation uses mathematical rules to compactly describe the semantics,
and it is used for proofs such as type soundness.
On the other hand, prose notation uses psudocodes algorithm to explain the
semantics through a step by step instruction.
Most WebAssembly users and engine developers are not familiar with
mathematical rules, so they utilize the prose notation.

% structure
A WebAssembly program is comprised of \textit{modules}.
A module is the unit of deployment, loading, and compilation, which consists of
definitions such as functions, globals, and memories.
A module is instantiated to validate the module and allocate each definition to
the data structure named \textit{store}.
This results in a module instance, a runtime representation for the module.
After the instantiation, functions can be invoked so that the function body,
which is a sequence of instructions, is executed.

% execution
WebAssembly execution is based on a stack machine.
An instruction consumes values in the stack as operands, performs some
operation, produces resulting values, and pushes them to the stack.
\cref{fig:testop} is the WebAssembly specification document for the
\texttt{testop} instruction.
Prose notation is written in the upper part, and formal notation is written in
the lower grey box.
In the prose notation, it pops a value from the stack, performs a test
operation, and pushes the result to the stack.
It is written compactly in the form of reduction rule in the formal notation
which consists of LHS, RHS, and premises: $(t.const ~ c_1) ~ t.testop$,
$(i32.const ~ c)$, and $c = testop_t(c_1)$.
The LHS means that the value at the top of the stack is $(t.const ~ c_1)$,
and an instruction to execute is $t.testop$.
The RHS means that the value at the top of the stack is changed to $(i32.const
~ c)$, and this happens when the premise $c = testop_t(c_1)$ is satisfied.

% control structure
The stack stores not only the input and output values of the instructions but also
structures related to the control flow.
A \textit{call frame} and a \textit{label} are used for function call and
branching, respectively.
In exception handling proposal, there is a \textit{handler} for exception
handling, and other strucutres could be introduced in the future.
How these control structures are used for the control flow will be discussed in
\cref{sec:control flow in official prose}.

>>>>

@@

WebAssembly (Wasm) is a portable, binary instruction format designed for safe
and efficient execution across diverse environments.
Initially developed to enhance performance in web browsers, diverse ecosystems
leverages the advantages of Wasm, including cloud and edge
computing~\cite{lucet, cloudflare}, the Internet of Things (IoT)\cite{wasm-iot}, and
blockchain technologies\cite{wasm-blockchain}.
Wasm programs are compiled from high-level languages (e.g., C, C++, Rust) into
a compact, platform-independent binary format, enabling consistent execution
across different hardware and operating systems.

% Structure
A Wasm program is comprised of \textit{modules}.
A module is the unit of deployment, loading, and compilation, which contains
definitions such as functions, tables, memories, and globals.
Functions represent units of executable code.
Each function has a type specifying its parameter and result types, along with
a body composed of Wasm instructions.
Tables represent indexed collections of references, such as function
references.
Tables are commonly used for indirect function calls, enabling features like
dynamic dispatch.
Memories represent linear memory, a contiguous, resizable array of bytes that
serves as the module’s heap.
Memory is used for storing and accessing data such as arrays, strings, or other
data structures.
Globals represent mutable or immutable variables that hold single values.
Globals can store values such as integers or floating-point numbers and are
accessible throughout the module.

% Instantiation & store
Before executing the code in a module, the module must be instantiated to
validate its content and allocate its definitions into a runtime data structure
called the \textit{store}.
The store acts as the memory space where executable code for functions are
stored, where function references are placed in tables, where the linear memory
is allocated for data, and where global variables are maintained.
Once instantiated, functions in the module can be invoked.
Invoking a function triggers the execution of its body, a sequence of
Wasm instructions.
These instructions operate on the module’s internal state, including the values
stored in the store.

% execution
Wasm execution is based on a stack machine.
An instruction consumes values from the stack as operands, performs an
operation, and pushes the resulting values onto the stack.
For example, the \texttt{testop} instruction exemplifies how this execution
model operates in practice.
\cref{fig:testop} illustrates the Wasm specification document for the
\texttt{testop} instruction.
It describes its semantics using two complementary forms: prose notation and
formal notation.
The prose notation is written in the upper part, and the formal notation is
written in the lower grey box.
In the prose notation, the instruction pops a value from the stack, performs a
test operation, and pushes the result back onto the stack.
In the formal notation, this is compactly expressed as a reduction rule
consisting of the left-hand side (LHS), right-hand side (RHS), and premises:
$(t.const ~ c_1) ~ t.testop$, $(i32.const ~ c)$, and $c = testop_t(c_1)$,
respectively.
The LHS indicates that the value at the top of the stack is $(t.const ~ c_1)$,
and the instruction to execute is $t.testop$.
The RHS shows that the value at the top of the stack is updated to $(i32.const
~ c)$, which occurs if the premise $c = testop_t(c_1)$ holds true.
% Additionally, the stack also stores structures related to the control flow
% along with the input and output values of the instructions: A \textit{frame}
% for function calls and a \textit{label} for branching.

### 2024-11-21 paper/dsl.tex
Write background

\chapter{Overview}
\label{ch:overview}
\noindent

\begin{figure}[t]
  \centerline{\includegraphics[width=15cm]{fig/overview}}
  \caption[An overview of the SpecTec architecture]
    {An overview of the SpecTec architecture}
    \label{fig:overview}
\end{figure}

% Overview explanation
In this chapter, we explain the overall approach to generating multiple
artifacts from the specification and ensuring the correctness of the
specification.
\Cref{fig:overview} illustrates an overview of our approach.
% SpecTec: specification mechanization tool
SpecTec is a tool for \textit{mechanizing} WebAssembly specification
~\cite{spectec}.
Specification mechanization is a technique that treats a specification as data
that can be manipulated by a machine to automatically generate parser and
interpreter ~\cite{jiset}, to generate tests and perform differential tests
~\cite{jest}, to find meta-level errors in the specification ~\cite{jstar}, and
to perform meta-level static analysis for the static analysis of the defined
language ~\cite{jsaver}.
This technique is especially crucial in growing languages such as JavaScript
and WebAssembly.

% DSL
SpecTec provides a domain-specific language (DSL) for writing specifications.
It is a declarative language designed to resemble the formal notation while
offering a compact and user-friendly syntax.
The code below shows the specification for the \texttt{testop} instruction
written in the DSL:
\begin{lstlisting}[escapechar=\%]
  rule Step_pure/testop:
  (CONST nt c_1) (TESTOP nt testop)  %$\tildecode$%>  (CONST I32 c)
  -- if c = %$\dollarcode$%testop_(nt, testop, c_1)
\end{lstlisting}
The DSL is type checked to prevent meta-level errors such as notation misuses and
dimension mismatches.
Using the DSL, SpecTec generates LaTeX code and mechanized definitions for theorem provers.

% AL
Additionally, SpecTec introduces another imperative language, \textit{AL}, for
the prose notation.
In AL, semantics are expressed algorithmically, in the form of step-by-step
instructions that take inputs and perform operations, mirroring the prose notation.
SpecTec translates declarative definitions in the DSL into algorithmic
definitions in AL.
The AL code below is the specification for the \texttt{testop} instruction
generated from the DSL:
\begin{lstlisting}[escapechar=\%]
Step_pure/testop nt testop {
 Assert (top_value(nt))
 Pop (numtype_0.CONST c_1)
 Let c = %$\dollarcode$%testop_(nt, testop, c_1)
 Push (I32.CONST c)
}
\end{lstlisting}
Using AL, SpecTec generates reStructuredText code.
\cref{fig:spectec-testop} shows the generated document for \texttt{testop}
instruction.
The generated document closely resembles the official document except for some
minor notation changes and some missing hyperlinks.

% AL interpreter
A notable feature of AL is that it is interpretable.
SpecTec includes an AL interpreter that executes an AL program with given inputs
and produces results.
Since an AL program specify the behavior of a Wasm program, the input to an AL
program is a Wasm program, and the result is the execution outcome of the Wasm
program.
In other words, the generated AL program functions as a WebAssembly interpreter.
What is particularly intriguing is that if specification authors write a Wasm
specification using the DSL, a WebAssembly interpreter that adheres to the
defined semantics is automatically generated.
By testing this generated interpreter, we can assess the specification,
thereby ensuring its correctness.

\red{(Explanation for each chapter for SpecTec)}

>>>>

\chapter{DSL}
\label{ch:dsl}
\noindent

### 2024-11-21 paper/model.tex
Write background

% !TEX root = main.tex

\chapter{Motivation}
\label{ch:motivation}
\noindent

\newcommand{\officialp}{official prose}
\newcommand{\spectecp}{SpecTec prose}

\red{TODO: explanation of the terms \officialp{} and \spectecp{}} \\
\red{TODO: remove redundant part in the fig} \\
\red{TODO: explanation of context(frame/label): in the background? or here?} \\
\red{TODO: the notion of meta-level interpreter in background} \\
\red{TODO: try finding better terminology rather than "model"} \\

\section{Control flow in official prose}
\label{sec:control flow in official prose}

% control flow structure in official prose
It might be easy to understand how the \officialp{} explains the control flow
of WebAssembly, if we assume that a WebAssembly code is loaded on a memory and
a pc points to the instruction to execute.
To describe control flow, it uses a structure named \textit{block} which
consitutes of a instruction sequence.
When executing the instructions in the block, pc can be changed to the starting
point of the block or the end of the block.

@@

\cref{ex:br} is a WebAssembly code example that has \texttt{loop} and
\texttt{unreachable}
Here, result type of the \texttt{loop} is \texttt{i32} and the \texttt{loop}
has a block of three instructions: \texttt{i32.const}, \texttt{br}, and
\texttt{unreachable}.
\cref{fig:loop} is the \officialp{} of the \texttt{loop} instruction.
It says that the continuation is the start of the loop, the information is
stored in a label, and it \textit{enters} the block with the label.
The term \textit{enter} means that it pushes the label in the stack and makes
pc points to the first instruction in the block.
The \texttt{i32.const} instruction is executed first, which just pushes the
\texttt{i32} value \texttt{42} in the stack.
Then, The \texttt{br} instruction is executed.
\cref{fig:br} is the \officialp{} of the \texttt{br} instruction.
It pops the label from the stack, and makes pc points to the continuation of
the label, which is the start of the loop instruction.
As a result, the code example above is a infinite loop pushing \texttt{42}
forever without executing \texttt{unreachable} instructions.

@@

% exiting label in official prose
There is also a special form of a behavior related to the block: an
\textit{exiting label}.
\cref{fig:exiting-label} is the \officialp{} of the \textit{exiting label}.
It is special because this behavior is performed without an explicit
WebAssembly instruction.
Rather, the behavior is performed when \textbf{the end of a block is reached}
without control instructions or runtime error.
The behavior is that the label is popped from the stack, and the pc changes to
the point after the end of the block.

@@

\cref{ex:exit} is a WebAssembly code example that has \texttt{loop} whose block is
only \texttt{i32.const}, and \texttt{f32.const}
When the \texttt{loop} instruction is executed, it enters the block with a label.
After \texttt{i32.const} is executed, the end of the block is reached.
As a result, exiting label occurs so that the label is popped from the stack,
pc changes to the point after the end of block: \texttt{f64.const}.
As a result, after executing the code, the value 42 and 3.14 pushed to the
stack.

>>>>

\chapter{Model}
\label{ch:Model}
\noindent

% TODO: Remove these commands
\newcommand{\officialp}{official prose}
\newcommand{\spectecp}{SpecTec prose}

% control structure
It might be easy to understand how the control flow of Wasm works, if we assume
that a Wasm code is loaded on a memory space and a pc points to the instruction
to execute.
To describe control flow, it introduces a structure named \textit{block} which
consitutes of a instruction sequence.
When branching instruction in the block is performed, pc can be changed to the
starting point of the block or the end of the block.

@@

\cref{ex:br} is a Wasm code example that has \texttt{loop} and
\texttt{unreachable}.
Here, result type of the \texttt{loop} is \texttt{i32} and the \texttt{loop}
has a block of three instructions: \texttt{i32.const}, \texttt{br}, and
\texttt{unreachable}.
\cref{fig:loop} is the specification for the \texttt{loop} instruction in the
prose notation.
In forth step, it says that the continuation is the start of the loop, the
information is stored in a label.
In the last step, it \textit{enters} the block with the label.
The term \textit{enter} means that it pushes the label onto the stack and makes
pc points to the first instruction in the block.
The \texttt{i32.const} instruction is executed first, which just pushes the
\texttt{i32} value \texttt{42} in the stack.
Then, The \texttt{br} instruction is executed.
\cref{fig:br} is the \officialp{} of the \texttt{br} instruction.
It pops the label from the stack, and makes pc points to the continuation of
the label, which is the start of the loop instruction.
As a result, the code example above is a infinite loop pushing \texttt{42}
forever without executing \texttt{unreachable} instructions.

@@

% exiting label in official prose
There is also a special form of a behavior related to the block: an
\textit{exiting label}.
\cref{fig:exiting-label} is the \officialp{} of the \textit{exiting label}.
It is special because this behavior is performed without an explicit
Wasm instruction.
Rather, the behavior is performed when \textbf{the end of a block is reached}
without control instructions or runtime error.
The behavior is that the label is popped from the stack, and the pc changes to
the point after the end of the block.

@@

\cref{ex:exit} is a Wasm code example that has \texttt{loop} whose block is
only \texttt{i32.const}, and \texttt{f32.const}
When the \texttt{loop} instruction is executed, it enters the block with a label.
After \texttt{i32.const} is executed, the end of the block is reached.
As a result, exiting label occurs so that the label is popped from the stack,
pc changes to the point after the end of block: \texttt{f64.const}.
As a result, after executing the code, the value 42 and 3.14 pushed to the
stack.
